### 2018-03-13 submission/manuscript.Rmd
update with pat's suggestions for intro

Colorectal cancer (CRC) is a growing world-wide health problem in which the microbiota has been hypothesized to have a role in disease progression [@siegel_cancer_2016; @flynn_metabolic_2016]. Numerous studies using murine models of CRC have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in tumorigenesis. Numerous case-control studies have characterized the microbiota of individuals with colonic adenomas and carcinomas in an attempt to identify biomarkers of disease progression [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017]. Because current CRC screening recommendations are poorly adhered to due to socioeconomic status, test invasiveness, and frequency of tests, development and validation of microbiome-associated biomarkers for CRC progression could further attempts to develop non-invasive diagnostics [@GimenoGarca2012].

Recently, there has been an intense focus on identifying microbiota-based biomarkers yielding a seemingly endless number of candidate taxa. Some studies point towards mouth-associated genera such as Fusobacterium, Peptostreptococcus, Parvimonas, and Porphyromonas that are enriched in people with carcinomas [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017]. Other studies have identified members of *Akkermansia*, *Bacteroides*, *Enterococcus*, *Escherichia*, *Klebsiella*, *Mogibacterium*, *Streptococcus*, and *Providencia* are also associated with carciomas  [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012]. Additionally, *Roseburia* has been found in some studies to be more abundant in people with tumors but in other studies it has been found to be either less abundant or no different than what is found in subjects with normal colons  [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014]. There are strong results from tissue culture and murine models that *Fusobacterium nucleatum*, pks-positive strains of *Escherichia coli*, *Streptococcus gallolyticus*, and an entertoxin-producing strain of *Bacteroides fragilis* are important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017; @weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014]. These results point to a causative role for the microbiota in CRC pathogenesis as well as their potential as diagnostic biomarkers.

Most studies have focused on identifying biomarkers in patients with carcinomas but there is a greater clinical need to identify biomarkers associated with adenomas. Studies focusing on broad scale community metrics have found that measures such as the total number of Operational Taxonomic Units (OTUs) are decreased in those with adenomas versus controls [@sanapareddy_increased_2012]. Other studies have identified *Acidovorax*, *Bilophila*, *Cloacibacterium*, *Desulfovibrio*, *Helicobacter*, *Lactobacillus*, *Lactococcus*, *Mogibacterium*, and *Pseudomonas* to be enriched in those with adenomas [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017]. There are few genera that are enriched in patients with adenoma or carcinoma tumors.

Confirming some of these previous findings, a recent meta-analysis found that 16S rRNA gene sequences from members of  *Akkermansia*, *Fusobacterium*, and *Parvimonas* were fecal biomarkers for the presence of carcinomas [@shah_leveraging_2017]. Contrary to previous studies they found sequences similar to members of *Lactobacillus* and *Ruminococcus* to be enriched in patients with adenoma or carcinoma relative to those with normal colons [@baxter_microbiota-based_2016; @burns_virulence_2015; @zeller_potential_2014]. In addition, they found 16S rRNA gene sequences from members of *Haemophilus*, *Methanosphaera*, *Prevotella*, *Succinovibrio* were enriched in patients with adenoma and *Pantoea* were enriched in patients with carcinomas. Although this meta-analysis was helpful for distilling a large number of possible biomarkers, the aggregate number of samples included in the analysis (n = 509) was smaller than several larger case-control studies that have since been published [@baxter_microbiota-based_2016; @hale_shifts_2017]

Here we provide an updated meta-analysis using 16S rRNA gene sequence data from both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) from 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2]. We expand both the breadth and scope of the previous meta-analysis to investigate whether biomarkers describing the bacterial community or specific members of the community can more accurately classify patients as having adenoma or carcinoma. Our results suggest that the bacterial community changes as disease severity worsens and that a subset of the microbial community can be used to diagnose the presence of carcinoma.

>>>>

Colorectal cancer (CRC) is a growing world-wide health problem in which the microbiota has been hypothesized to have a role in disease progression [@siegel_cancer_2016; @flynn_metabolic_2016]. Numerous studies using murine models of CRC have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in tumorigenesis. Numerous case-control studies have characterized the microbiota of individuals with colonic adenomas and carcinomas in an attempt to identify biomarkers of disease progression [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017]. Because current CRC screening recommendations are poorly adhered to due to person's socioeconomic status, test invasiveness, and frequency of tests, development and validation of microbiota-associated biomarkers for CRC progression could further attempts to develop non-invasive diagnostics [@GimenoGarca2012].

Recently, there has been an intense focus on identifying microbiota-based biomarkers yielding a seemingly endless number of candidate taxa. Some studies point towards mouth-associated genera such as *Fusobacterium*, *Peptostreptococcus*, *Parvimonas*, and *Porphyromonas* that are enriched in people with carcinomas [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017]. Other studies have identified members of *Akkermansia*, *Bacteroides*, *Enterococcus*, *Escherichia*, *Klebsiella*, *Mogibacterium*, *Streptococcus*, and *Providencia* [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012]. Additionally, *Roseburia* has been found in some studies to be more abundant in people with tumors but in other studies it has been found to be less abundant than what is found in subjects with normal colons  [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014]. There is support from mechanistic studies using tissue culture and murine models that *Fusobacterium nucleatum*, pks-positive strains of *Escherichia coli*, *Streptococcus gallolyticus*, and an entertoxin-producing strain of *Bacteroides fragilis* are important in tumorigenesis [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017; @weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014]. These results point to a causative role for the microbiota in tumorigenesis as well as their potential as diagnostic biomarkers.

Most studies have focused on identifying biomarkers in patients with carcinomas but there is a clinical need to identify biomarkers associated with adenomas to facilitate early detection of the tumors. Studies focusing on broad scale community metrics have found that measures such as the total number of taxa (i.e. richness) are lower in those with adenomas versus controls [@sanapareddy_increased_2012]. Other studies have identified *Acidovorax*, *Bilophila*, *Cloacibacterium*, *Desulfovibrio*, *Helicobacter*, *Lactobacillus*, *Lactococcus*, *Mogibacterium*, and *Pseudomonas* to be enriched in those with adenomas [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017]. The ability to classify individuals as having normal colons or adenomas based solely on the taxa within fecal samples has been limited. However, when 16S rRNA gene sequence data was combined with the results of a fecal immunochemical test (FIT), the ability to diagnose individuals with adenomas was improved relative to using the FIT results alone [@baxter_microbiota-based_2016].

A recent meta-analysis found that 16S rRNA gene sequences from members of  *Akkermansia*, *Fusobacterium*, and *Parvimonas* were fecal biomarkers for the presence of carcinomas [@shah_leveraging_2017]. Contrary to previous studies they found sequences similar to members of *Lactobacillus* and *Ruminococcus* to be enriched in patients with adenoma or carcinoma relative to those with normal colons [@baxter_microbiota-based_2016; @burns_virulence_2015; @zeller_potential_2014]. In addition, they found that 16S rRNA gene sequences from members of *Haemophilus*, *Methanosphaera*, *Prevotella*, and *Succinovibrio* were enriched in patients with adenomas and *Pantoea* were enriched in patients with carcinomas. Although this meta-analysis was helpful for distilling a large number of possible biomarkers, the aggregate number of samples included in the analysis (n=509) was smaller than several larger case-control studies that have since been published [@baxter_microbiota-based_2016; @hale_shifts_2017]

Here we provide an updated meta-analysis using 16S rRNA gene sequence data from both feces (n=`r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) from 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2]. We expand both the breadth and scope of the previous meta-analysis to investigate whether biomarkers describing the bacterial community or specific members of the community can more accurately classify patients as having adenoma or carcinoma. Our results suggest that the bacterial community changes as disease severity worsens and that a subset of the microbial community can be used to diagnose the presence of carcinoma.
